---
title: "Basic model of evolutionary rescue in discrete time"
author: "Iona & Pascal"
date: "2025-10-15"


output:
  html_document:
    df_print: paged
---

## Simulating evolutionary rescue in discrete time

In this notebook, we are implementing simulations of evolutionary rescue by introducing an other species. The model is as follows: a diploid homozygot population (aa) is dying of, due to a recessive fitness disadvantage on a. At time x a few homozygot individuals (AA) are introduced to save the population by hybridization.

**Idea:** Inbreed depression in population with recessive disease (aa) that leads to lower fitness and is therefore declining in population size. Goal is to rescue population with hybrids. How many healthy individuals are needed to rescue the population. only healthy (AA) are added,

**Assumptions:**

\- starting population is homozygot for the disease (aa)

\- healthy added individuals are homozygot without the disease (AA)

\- second generation can mate with sick individual or hybrid

**Question:** How many individuals have to be introduced at time x to rescue a homozygot population with low fitness (inbreed depression)?

=======

First we define the variables and parameters.

*Parameters we need:*

-   `N_aa` initial population size of the genotype aa
-   `N_aA` initial population size of the genotype aA
-   `N_AA` initial population size of the genotype AA

-   `r_aa` fitness of genotype aa
-   `s_aA` fitness of genotype aA
-   `s_AA` fitness of genotype AA

-   `max_gen`

*Variables we need:*

-   `offsp_aa` number of offspring of the given genotype
-   `N_aa_new` number of the given genotype after one generation
-   `total_gametes` number of all alleles combined
-   `freq_a` number of the given allele in the population
-   `predicted_freq_aa` number of the given genotype in the next generation
-   `t_max` number of generation till extinction/ rescue

*Functions we need:*

-   `simulate_gen` models one generation step




## Simulate one generation of population size change


```{r}
N_aa <- 100
N_aA <- 0
N_AA <- 3


r_aa <- 0.2 # fitness reduction of aa allel comb. equal to decay_rate
s_aA <- 1.10 # fitness benefit of allel A, equal to sel_coeff
s_AA <- 1.10 # fitness benefit of allel A, equal to sel_coeff

# --- Simulation of one generation with three genotypes ---
simulate_gen <- function(N_aa, N_aA, N_AA, r_aa, s_aA, s_AA) {

  # Reproduction with fitness malus/benefit
  offsp_aa <- rpois(1, N_aa * (1-r_aa))
  offsp_aA <- rpois(1, N_aA * s_aA)
  offsp_AA <- rpois(1, N_AA * s_AA)  
  
  # Total numbeer of gamets proportional to the frequency of alleles A and a
  total_gametes <- 2 * (offsp_aa + offsp_aA + offsp_AA)
  freq_A <- (2 * offsp_AA + offsp_aA) 
  freq_a <- total_gametes - freq_A
  
  if (total_gametes > 0) {

  
  #print(freq_a)
  #print(total_gametes)
  
  # Change of allele frequencies (A -> a und a -> A); p = (freq_A / total_gametes), q = (freq_a / total_gametes) #hardy equilibrium 1 = p^2 +2pq +q^2
  predicted_freq_aa <- (freq_a/total_gametes)^2
  predicted_freq_aA <- 2*(freq_A / total_gametes)*(freq_a / total_gametes)
  predicted_freq_AA <- (freq_A/total_gametes)^2
  
  #print(predicted_freq_aA)
  #print("PROBLEM")
 
   # Random draw of new genotypes under hardy-weinberg conditions
  N_total <- offsp_aa + offsp_aA + offsp_AA
  
  #print(c(predicted_freq_aa, predicted_freq_aA, predicted_freq_AA))
  #print(N_total)
  
  genotypes <- rmultinom(1, N_total, prob = c(predicted_freq_aa, predicted_freq_aA, predicted_freq_AA))
  
  #print(genotypes)

  
  N_aa_new <- genotypes[1]
  N_aA_new <- genotypes[2]
  N_AA_new <- genotypes[3]
  
  #print(c(N_aa_new, N_AA_new, N_aA_new)) 
  return(c(N_aa_new, N_aA_new, N_AA_new))
  }
  
  else
  return(c(0,0,0))
  
  
}

print(simulate_gen(N_aa, N_aA, N_AA, r_aa, s_aA, s_AA))
```

## Simulate a population trajectory

```{r}
simulate_pop <- function(N_aa, N_aA, N_AA, r_aa, s_aA, s_AA, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_aa, N_aA, N_AA)
  # initiate the variables
  pop_new <- c(N_aa, N_aA, N_AA)
  
  # run the simulation until generation t_max
  for (i in 1:(t_max+1)) {
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]+pop_new[3]>=100*(N_aa+N_aA+N_AA) | pop_new[1]+pop_new[2]+pop_new[3]==0) break
    #print(pop_new[1]+pop_new[2]+pop_new[3])
    # redefine the current population one generation later
    pop_new <- simulate_gen(pop_new[1],pop_new[2],pop_new[3], r_aa, s_aA, s_AA)
    #print(pop_new)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("aa","aA","AA")
  # return the result
  #print(pop_vector)
  return(pop_vector)	
}

# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 1000
# create your simulation data
output <- simulate_pop(N_aa, N_aA, N_AA, r_aa, s_aA, s_AA, max_gen)
# show the last few lines of the data table
print(tail(output))

### Next time need to plot and see if distribution is right?

# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2]+output[,3],type='l',ylim=c(0,max(output[,1]+output[,2]+output[,3])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")  #aa
# add number of mutant individuals
lines(x_range,output[,2], col="red")  #aA
# add number of mutant individuals
lines(x_range,output[,3], col="green")  #AA

```



```{r}
# set some parameters to fixed values
N_aA <- 0

r_aa <- 0.2
s_AA <- 1.1


t_max <- 10
# determine how often to run the simulation for each set of parameters
no_replicates <- 10

# set parameters to vary
N_aa <- c(10,100,50,40)
N_AA <- c(2,5,7,10)
s_aA <- c(1.1,1.2,1,0.9) 

# initialize data table - where to collect the results
data_table <- c()
# run the simulation across all chosen parameters
# loop over decay rates
for(Naa in N_aa){
  # loop over selection coefficients
  for(NAA in N_AA){
    for(saA in s_aA){
       # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
      i<-1
      repeat {
        # increase counter by one
        i<-i+1
        # run the simulation once
        one_run <- simulate_pop(N_aa, N_aA, N_AA, r_aa, s_aA, s_AA, t_max)
        # determine total population sizes
        total_size <- one_run[,1]+one_run[,2]
        # determine minimum population size
        min_size <- min(total_size)
        # determine (first) generation at which this population size occurred
        min_gen <- as.numeric(which(total_size==min_size)[1])
        # enter the data into the table
        data_table <- rbind(data_table,c(rval,sval,min_gen,min_size)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
        # stop the repeated computation after no_replicates times
        if(i>no_replicates) break
        }
      }
  }
}
# define column names of data table
colnames(data_table) <- c("r","s","min_gen","no_min")
# show the first lines of the output
print(head(data_table))
```

